### 2018-01-19 derivations.tex
edits

\newcommand{\pluseq}{\mathrel{{+}{=}}}
\newcommand{\minuseq}{\mathrel{{-}{=}}}
\newcommand{\ns}{N_{sp}}
\newcommand{\nr}{N_{reac}}
\newcommand{\conp}{CONP}
\newcommand{\conv}{CONV}
\newcommand{\dconp}{\ifmmode\text{for \conp,}\else for \conp,\fi}
\newcommand{\dconv}{\ifmmode\text{for \conv,}\else for \conv,\fi}
\newcommand{\Ru}{\mathcal{R}}

@@

\subsection{Thermo-chemical source terms}
The evolution of the thermo-chemical state variables of this system is described by a set of ordinary-differential equations:
\begin{subequations}
\label{e:source_terms}
\begin{align}
f &= \frac{\text{d} \Phi }{\text{d} t } = \left\{\frac{\text{d} T }{\text{d} t },\frac{\text{d} V }{\text{d} t },\frac{\text{d} n_1}{\text{d} t },\frac{\text{d} n_2 }{\text{d} t }\ldots \frac{\text{d} n_{\ns - 1} }{\text{d} t }\right\}\dconp, \\
f &= \frac{\text{d} \Phi }{\text{d} t } = \left\{\frac{\text{d} T }{\text{d} t },\frac{\text{d} P }{\text{d} t },\frac{\text{d} n_1}{\text{d} t },\frac{\text{d} n_2 }{\text{d} t }\ldots \frac{\text{d} n_{\ns - 1} }{\text{d} t }\right\}\dconv
\end{align}
\end{subequations}
For both, the molar source terms are~\cite{TurnsStephenR2012Aitc}:
\begin{equation}
\frac{\text{d} n }{\text{d} t }_{k} = V \dot{\omega}_{k} \label{e:spec}
\end{equation}
where $\dot{\omega}_k$ is the $kth$ species overall production rate, and the temperature production rates are~\cite{TurnsStephenR2012Aitc}:
\begin{subequations}
\label{e:temperature_incomplete}
\begin{align}
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns} H_{k} \dot{\omega}_{k}}{\sum_{k=1}^{\ns} [C]_{k} {C_{p, k}}} & \text{\dconp} \\
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns} U_{k} \dot{\omega}_{k}}{\sum_{k=1}^{\ns} [C]_{k} {C_{v, k}}} & \text{\dconv}
\end{align}
\end{subequations}
where $H_k$, $U_k$, $C_{p,k}$ and $C_{v, k}$ are the enthalpy, internal energy, constant-pressure specific heat and constant-volume specific heat of species $k$ in molar units respectively, while $[C]_{k}$ is the concentration, given by:
\begin{equation}
 [C]_{k} = \frac{n_{k}}{V}
\end{equation}

@@

Combining~\cref{e:spec,e:param_incomplete,e:total_molar} gives the final form of the pressure and volume terms:
\begin{subequations}
\label{e:param_complete}
\begin{align}
\frac{\text{d} V }{\text{d} t } &= V \left(\frac{T \Ru}{P} \sum_{k=1}^{-1 + \ns} \left(1 - \frac{W_{k}}{W_{\ns}}\right) \dot{\omega}_{k} + \frac{1}{T} \frac{\text{d} T }{\text{d} t }\right)& \text{\dconp} \\
\frac{\text{d} P }{\text{d} t } &= \frac{P}{T} \frac{\text{d} T }{\text{d} t } + T \Ru \sum_{k=1}^{-1 + \ns} \left(1 - \frac{W_{k}}{W_{\ns}}\right) \dot{\omega}_{k}&\text{\dconv}
\end{align}
\end{subequations}

Additionally, the concentration and moles of the last species in the model (which is not a state variable) may be expanded in terms of the state variables of the system:
\begin{equation}
 \label{e:last_spec_conc}
   [C]_{\ns} = [C] - \sum_{k=1}^{-1 + \ns} [C]_{k}
\end{equation}
where $[C]$ is the total concentration:
\begin{equation}
 [C] = \frac{P}{T \Ru}
\end{equation}

Using~\cref{e:last_spec_conc,e:spec,e:spec_ns}, the temperature source terms may be expanded to contain only state variables (i.e., by removal of the last species' source term and overall production rate):
\begin{subequations}
\label{e:temperature_complete}
\begin{align}
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{-1 + \ns} \left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k}}{[C] {C_{p,\ns}} + \sum_{k=1}^{-1 + \ns} \left(- {C_{p,\ns}} + {C_{p, k}}\right) [C]_{k}} &\text{\dconp}\\
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{-1 + \ns} \left(U_{k} - \frac{W_{k} U_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k}}{[C] {C_{v,\ns}} + \sum_{k=1}^{-1 + \ns} \left(- {C_{v,\ns}} + {C_{v, k}}\right) [C]_{k}} &\text{\dconv}
\end{align}
\end{subequations}
This form is used for differentiation, but~\cref{e:temperature_incomplete} will often be used as well due to its compactness.

@@

For a reaction enhanced (or diminished) by the presence of a third body, the reaction rate is modified by the third-body concentration $[X]_i$ given by
\begin{equation}
[X]_{i} = \sum_{k=1}^{\ns} \alpha_{k,i} [C]_{k} \;,
\end{equation}
where $\alpha_{k,i}$ is the third-body efficiency of species $k$ in the $i$th reaction.
For a default third-body efficiency of $\alpha_{k,i} = 1$, this may be rearranged to the compact-storage form:
\begin{equation}
 [X]_{i} = [C] + \sum_{k=1}^{\ns} \left(\alpha_{k,i} - 1\right) [C]_{k} \;,
\end{equation}
where only species with non-default third-body efficiencies must be stored in the model.
Expanding the concentration of the last species gives:
\begin{equation}
\label{e:thd_mix}
 [X]_{i}=[C] \alpha_{\ns,i} + \sum_{k=1}^{-1 + \ns} \left(- \alpha_{\ns,i} + \alpha_{k,i}\right) [C]_{k}\;.
\end{equation}
This form will be denoted as the mixture-based (or \textbf{mix} for short) third-body efficiency.

@@

In addition, a single species may act as the third body in which case
\begin{equation}
 [X]_{i} = [C_m]
\end{equation}
where the $m$th species is the third body.
For evaluation of Jacobian entries, the expanded form:
\begin{equation}
\label{e:thd_spec}
 [X]_{i}=\left([C] - \sum_{k=1}^{-1 + \ns} [C]_{k}\right) \delta_{\ns m} + \left(- \delta_{\ns m} + 1\right) [C]_{m}
\end{equation}
will be used, where the Kronecker delta $\delta_{\ns m}$ is unity if and only if the third body species $m$ is the last species in the model.
\Cref{e:thd_spec} will be referred to as the \textbf{species}-based third-body efficiency.

@@

Let $\mathcal{J}$ denote the Jacobian matrix corresponding to the set of ODEs defined in~\cref{e:source_terms}.
$\mathcal{J}$ is filled with the partial derivatives $\partial f / \partial \Phi$, such that:
\begin{equation}
 \mathcal{J}_{i,j} = \frac{\partial f_i}{\partial \Phi_j},\qquad j=1 \ldots \ns - 1
\end{equation}
We will begin by deriving subcomponents of the Jacobian matrix in the subsequent sections before presenting the final form of the Jacobian matrix in~\cref{s:jac_final}.

@@

\subsection{Rate of progress derivatives}
\subsubsection{Temperature derivative}
\label{s:dri_dt}
Next, the rate of progress derivatives are evaluated, first with respect to temperature:
\begin{equation}
 \frac{\partial R }{\partial T }_{i} = \frac{\partial}{\partial T}\left({k_f}_{i} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime}_{k,i}}\right)
\end{equation}
The derivative of the forward reaction rate coefficient is:
\begin{equation}
 \label{e:dkf_dt}
 \frac{\text{d} {k_f} }{\text{d} T }_{i} = \frac{{k_f}_{i}}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)
\end{equation}
While evaluating derivatives for the Jacobian, it is important to expand terms---e.g., concentration, production rate, etc.---related to the last species to obtain the correct derivative, in this case using \cref{e:last_spec_conc}:
\begin{equation}
 {R_f} = \left(\left(- \sum_{k=1}^{-1 + \ns} [C]_{k} + \frac{P}{T \Ru}\right)^{\nu^{\prime}_{\ns,i}}\right) {k_f}_{i} \prod_{k=1}^{-1 + \ns} [C]_{k}^{\nu^{\prime}_{k,i}}
\end{equation}
The derivative of a species concentration with respect to temperature is zero for both \conp~and \conv:
\begin{align}
 \frac{\partial [C]_{k}}{\partial T} &= \frac{\partial}{\partial T} \left(\frac{n_k}{V}\right), \quad \left\{k \ne \ns\right\} \nonumber \\
				     &= 0
\end{align}
as the volume is either constant (\conp) or a state variable and a function of time only (\conv).
Hence, the full derivative of the forward rate of progress is:
\begin{equation}
 \label{e:dropfdt_1}
 \frac{\partial {R_f} }{\partial T }_{i} = \frac{\text{d} {k_f} }{\text{d} T }_{i} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime}_{k,i}} - \frac{[C] \nu^{\prime}_{\ns,i}}{T} [C]_{\ns}^{\nu^{\prime}_{\ns,i} - 1} {k_f}_{i} \prod_{k=1}^{-1 + \ns} [C]_{k}^{\nu^{\prime}_{k,i}}
\end{equation}
By defining a temporary variable $S^{\prime}_{l}$ (for species $l = \left\{j, \ns\right\}$):
\begin{equation}
 \label{e:s_temp}
 S^{\prime}_{l} = \nu^{\prime}_{l,i} [C]_{l}^{\nu^{\prime}_{l,i} - 1} \prod_{\substack{1 \leq l \leq l - 1\\l + 1 \leq l \leq \ns}} [C]_{l}^{\nu^{\prime}_{l,i}}
\end{equation}
and using~\cref{e:dkf_dt}, \cref{e:dropfdt_1} can be simplified to:
\begin{equation}
 \label{e:dropf_dt}
 \frac{\partial {R_f} }{\partial T }_{i} = - \frac{[C] S^{\prime}_{\ns}}{T} {k_f}_{i} + \frac{{R_f}_{i}}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)
\end{equation}

Starting from~\cref{e:ropr} and applying the same expansion of the last species' concentration as previously, the temperature derivative of the reverse rate of progress is:
\begin{equation}
 \label{e:droprdt_1}
 \frac{\partial {R_r} }{\partial T }_{i} = \frac{\text{d} {k_r} }{\text{d} T }_{i} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime\prime}_{k,i}} - \frac{[C] \nu^{\prime\prime}_{\ns,i}}{T} [C]_{\ns}^{\nu^{\prime\prime}_{\ns,i} - 1} {k_f}_{i} \prod_{k=1}^{-1 + \ns} [C]_{k}^{\nu^{\prime\prime}_{k,i}}
\end{equation}
Next, \cref{e:kr} is considered to obtain the temperature derivative of the non-explicit reversible reaction rate coefficient:
\begin{align}
 \frac{\text{d} {k_r} }{\text{d} T }_{i} &= \left(- \frac{1}{{K_c}_{i}} \frac{\text{d} {K_c}_{i} }{\text{d} T } \frac{{k_f}_i}{{K_c}_i} + \frac{1}{{K_c}_{i}}\frac{\text{d} {k_f}_{i} }{\text{d} T } \right) \nonumber \\
				         &= \left(- \frac{1}{{K_c}_{i}} \frac{\text{d} {K_c}_{i} }{\text{d} T } + \frac{1}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)\right) {k_r}_{i}
\end{align}
The temperature derivative of the equilibrium constant is:
\begin{equation}
 \frac{\text{d} {K_c}_{i} }{\text{d} T } = {K_c}_{i} \sum_{k=1}^{\ns} \nu_{k,i} \frac{\text{d} B }{\text{d} T }_{k} \;,
\end{equation}
with:
\begin{equation}
 \frac{\text{d} B }{\text{d} T }_{k} = T \left(T \left(\frac{T a_{k,4}}{5} + \frac{a_{k,3}}{4}\right) + \frac{a_{k,2}}{3}\right) + \frac{a_{k,1}}{2} + \frac{1}{T} \left(a_{k,0} - 1 + \frac{a_{k,5}}{T}\right)
\end{equation}
giving:
\begin{equation}
\label{e:dkr_dt}
\frac{\text{d} {k_r} }{\text{d} T }_{i} = \left(- \sum_{k=1}^{\ns} \nu_{k,i} \frac{\text{d} B }{\text{d} T }_{k} + \frac{1}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)\right) {k_r}_{i} \;,
\end{equation}
and using an temporary value $S^{\prime\prime}_{\ns}$---defined analagously to \cref{e:s_temp} for the reverse direction---we obtain:
\begin{equation}
\label{e:dropr_dt}
\frac{\partial {R_r} }{\partial T }_{i} = \left(- \sum_{k=1}^{\ns} \nu_{k,i} \frac{\text{d} B }{\text{d} T }_{k} + \frac{1}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)\right) {R_r}_{i} - \frac{[C] S^{\prime\prime}_{\ns}}{T} {k_r}_{i}
\end{equation}

>>>>

\newcommand{\pluseq}{\mathrel{{+}{=}}}
\newcommand{\minuseq}{\mathrel{{-}{=}}}
\newcommand{\ns}{N_{sp}}
\newcommand{\nr}{N_{reac}}
\newcommand{\conp}{CONP}
\newcommand{\conv}{CONV}
\newcommand{\dconp}{\ifmmode\text{for \conp,}\else for \conp,\fi}
\newcommand{\dconv}{\ifmmode\text{for \conv,}\else for \conv.\fi}
\newcommand{\Ru}{\mathcal{R}}

@@

\subsection{Thermo-chemical source terms}
The evolution of the thermo-chemical state variables of this system is described by a set of ordinary-differential equations:
\begin{subequations}
\label{e:source_terms}
\begin{align}
f &= \frac{\text{d} \Phi }{\text{d} t } = \left\{\frac{\text{d} T }{\text{d} t },\frac{\text{d} V }{\text{d} t },\frac{\text{d} n_1}{\text{d} t },\frac{\text{d} n_2 }{\text{d} t }\ldots \frac{\text{d} n_{\ns - 1} }{\text{d} t }\right\}\dconp, \\
f &= \frac{\text{d} \Phi }{\text{d} t } = \left\{\frac{\text{d} T }{\text{d} t },\frac{\text{d} P }{\text{d} t },\frac{\text{d} n_1}{\text{d} t },\frac{\text{d} n_2 }{\text{d} t }\ldots \frac{\text{d} n_{\ns - 1} }{\text{d} t }\right\}\dconv
\end{align}
\end{subequations}
For both, the molar source terms are~\cite{TurnsStephenR2012Aitc}:
\begin{equation}
\frac{\text{d} n }{\text{d} t }_{k} = V \dot{\omega}_{k} \label{e:spec}
\end{equation}
where $\dot{\omega}_k$ is the $kth$ species overall production rate, and the temperature production rate~\cite{TurnsStephenR2012Aitc} is:
\begin{subequations}
\label{e:temperature_incomplete}
\begin{align}
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns} H_{k} \dot{\omega}_{k}}{\sum_{k=1}^{\ns} [C]_{k} {C_{p, k}}} & \text{\dconp} \\
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns} U_{k} \dot{\omega}_{k}}{\sum_{k=1}^{\ns} [C]_{k} {C_{v, k}}} & \text{\dconv}
\end{align}
\end{subequations}
where $H_k$, $U_k$, $C_{p,k}$ and $C_{v, k}$ are the enthalpy, internal energy, constant-pressure specific heat and constant-volume specific heat of species $k$ in molar units respectively, while $[C]_{k}$ is the concentration, given by:
\begin{equation}
 [C]_{k} = \frac{n_{k}}{V}
\end{equation}

@@

Combining~\cref{e:spec,e:param_incomplete,e:total_molar} gives the final form of the pressure and volume terms:
\begin{subequations}
\label{e:param_complete}
\begin{align}
\frac{\text{d} V }{\text{d} t } &= V \left(\frac{T \Ru}{P} \sum_{k=1}^{\ns  - 1} \left(1 - \frac{W_{k}}{W_{\ns}}\right) \dot{\omega}_{k} + \frac{1}{T} \frac{\text{d} T }{\text{d} t }\right)& \text{\dconp} \\
\frac{\text{d} P }{\text{d} t } &= \frac{P}{T} \frac{\text{d} T }{\text{d} t } + T \Ru \sum_{k=1}^{\ns  - 1} \left(1 - \frac{W_{k}}{W_{\ns}}\right) \dot{\omega}_{k}&\text{\dconv}
\end{align}
\end{subequations}

Additionally, the concentration and moles of the last species in the model (which is not a state variable) may be expanded in terms of the state variables of the system:
\begin{equation}
 \label{e:last_spec_conc}
   [C]_{\ns} = [C] - \sum_{k=1}^{\ns  - 1} [C]_{k}
\end{equation}
where $[C]$ is the total concentration:
\begin{equation}
 [C] = \frac{P}{T \Ru}
\end{equation}

Using~\cref{e:last_spec_conc,e:spec,e:spec_ns}, the temperature source terms may be expanded to contain only state variables (i.e., by removal of the last species' source term and overall production rate):
\begin{subequations}
\label{e:temperature_complete}
\begin{align}
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns  - 1} \left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k}}{[C] {C_{p,\ns}} + \sum_{k=1}^{\ns  - 1} \left(- {C_{p,\ns}} + {C_{p, k}}\right) [C]_{k}} &\text{\dconp}\\
\frac{\text{d} T }{\text{d} t } &= - \frac{\sum_{k=1}^{\ns  - 1} \left(U_{k} - \frac{W_{k} U_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k}}{[C] {C_{v,\ns}} + \sum_{k=1}^{\ns  - 1} \left(- {C_{v,\ns}} + {C_{v, k}}\right) [C]_{k}} &\text{\dconv}
\end{align}
\end{subequations}
This form is used for differentiation, but~\cref{e:temperature_incomplete} will often be used as well due to its compactness.

@@

For a reaction enhanced (or diminished) by the presence of a third body, the reaction rate is modified by the third-body concentration $[X]_i$ given by
\begin{equation}
[X]_{i} = \sum_{k=1}^{\ns} \alpha_{k,i} [C]_{k} \;,
\end{equation}
where $\alpha_{k,i}$ is the third-body efficiency of species $k$ in the $i$th reaction.
For a default third-body efficiency of $\alpha_{k,i} = 1$, this may be rearranged to the compact-storage form:
\begin{equation}
 [X]_{i} = [C] + \sum_{k=1}^{\ns} \left(\alpha_{k,i} - 1\right) [C]_{k} \;,
\end{equation}
where only species with non-default third-body efficiencies must be stored in the model.
Expanding the concentration of the last species gives:
\begin{equation}
\label{e:thd_mix}
 [X]_{i}=[C] \alpha_{\ns,i} + \sum_{k=1}^{\ns  - 1} \left(- \alpha_{\ns,i} + \alpha_{k,i}\right) [C]_{k}\;.
\end{equation}
This form will be denoted as the mixture-based (or \textbf{mix} for short) third-body efficiency.

@@

In addition, a single species may act as the third body in which case
\begin{equation}
 [X]_{i} = [C_m]
\end{equation}
where the $m$th species is the third body.
For evaluation of Jacobian entries, the expanded form:
\begin{equation}
\label{e:thd_spec}
 [X]_{i}=\left([C] - \sum_{k=1}^{\ns  - 1} [C]_{k}\right) \delta_{\ns m} + \left(- \delta_{\ns m} + 1\right) [C]_{m}
\end{equation}
will be used, where the Kronecker delta $\delta_{\ns m}$ is unity if and only if the third body species $m$ is the last species in the model.
\Cref{e:thd_spec} will be referred to as the \textbf{species}-based third-body efficiency.

@@

Let $\mathcal{J}$ denote the Jacobian matrix corresponding to the set of ODEs defined in~\cref{e:source_terms}.
$\mathcal{J}$ is filled with the partial derivatives $\partial f / \partial \Phi$, such that:
\begin{equation}
 \label{e:jac_general}
 \mathcal{J}_{i,j} = \frac{\partial f_i}{\partial \Phi_j},\qquad j=1 \ldots \ns - 1
\end{equation}
The Jacobian entries resulting from~\cref{e:jac_general} are derived in~\cref{s:energy_derivs}, while various subcomponents of the Jacobian are derived in~\cref{s:dri_dt,s:dri_dnj,s:dri_de,s:dri_pdep,s:dpmod_fall} and the final form of the Jacobian matrix is presented in~\cref{s:jac_final}.

We note that some parts of the Jacobian derivation---in particular derivatives of the energy equation---are quite complicated.
The aim of this document is to provide the reader with an overview of the derivation process; some intermediate steps are left out due to brevity and typesetting constraints.
For a complete derivation, the reader is directed to the output of the \href{https://github.com/arghdos/SPyJac-paper/blob/master/derivations/scripts/derivations.py}{derivations.py} script, from which this document was compiled.

\subsection{Energy equation derivatives}
\label{s:energy_derivs}
\subsubsection{Temperature derivative}
In this section, we will more closely follow the \conp~energy equation derivative with respect to temperature as the process is largely identical between \conp~and \conv.

First, the derivative of full form of the energy source term~\cref{e:temperature_complete} with respect to temperature will be considered.
To make the result more managable, the denominator of~\cref{e:temperature_complete} is first collapsed back into the form of~\cref{e:temperature_incomplete}, yielding for~\conp:
 \frac{\partial\dot{T}}{\partial{T}} &= - \frac{\left(- [C] \frac{\text{d} {C_p} }{\text{d} T }_{\ns} - \sum_{k=1}^{-1 + \ns} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{\ns} + \frac{\text{d} {C_p} }{\text{d} T }_{k}\right) [C]_{k} + \frac{[C] {C_p}_{\ns}}{T}\right)}{\left(\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}\right)^{2}} \times \nonumber \\
				     & \mathindent[largerindent] \sum_{k=1}^{-1 + \ns} \left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k} - \nonumber \\
				     & \mathindent \frac{\sum_{k=1}^{-1 + \ns} \left(\left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \frac{\partial \dot{\omega} }{\partial T }_{k} + \left(\frac{\text{d} H }{\text{d} T }_{k} - \frac{W_{k}}{W_{\ns}} \frac{\text{d} H }{\text{d} T }_{\ns}\right) \dot{\omega}_{k}\right)}{\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}}
Next, we may recognize that~\cref{e:temperature_incomplete} may be factored out of the first terms, yielding:
 \frac{\partial\dot{T}}{\partial{T}} &= \frac{1}{\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}} \Biggl[ \frac{\text{d} T }{\text{d} t } \times \nonumber \\
				     & \left( - [C] \frac{\text{d} {C_p} }{\text{d} T }_{\ns} - \sum_{k=1}^{-1 + \ns} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{\ns} + \frac{\text{d} {C_p} }{\text{d} T }_{k}\right) [C]_{k} + \frac{[C] {C_p}_{\ns}}{T} \right) - \nonumber \\
				     & \sum_{k=1}^{-1 + \ns} \left(\left(H_{k} - \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \frac{\partial \dot{\omega} }{\partial T }_{k} + \left(\frac{\text{d} H }{\text{d} T }_{k} - \frac{W_{k}}{W_{\ns}} \frac{\text{d} H }{\text{d} T }_{\ns}\right) \dot{\omega}_{k}\right) \Biggr]
Now, we invoke the identity:
\begin{equation}
 [C] = \sum_{k=1}^{\ns} [C]_{k}
\end{equation}
to simplify the specific heat terms:
 \frac{\partial\dot{T}}{\partial{T}} &= \frac{1}{\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}} \Biggl[ \frac{\text{d} T }{\text{d} t } \sum_{k=1}^{\ns} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{k} + \frac{{C_p}_{\ns}}{T}\right) [C]_{k} + \nonumber \\
				     & \sum_{k=1}^{-1 + \ns} \biggl(\left(- H_{k} + \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \frac{\partial \dot{\omega} }{\partial T }_{k} + \nonumber \\
				     & \mathindent[larger] \left(- \frac{\text{d} H }{\text{d} T }_{k} + \frac{W_{k}}{W_{\ns}} \frac{\text{d} H }{\text{d} T }_{\ns}\right) \dot{\omega}_{k}\biggr)\Biggr]
and finally, we may substitute in the specific heat for the temperature derivative of enthalpy, giving:
 \frac{\partial\dot{T}}{\partial{T}} &= \frac{1}{\sum_{k=1}^{\ns} [C]_{k} {C_p}_{k}} \Biggl[ \frac{\text{d} T }{\text{d} t } \sum_{k=1}^{\ns} \left(- \frac{\text{d} {C_p} }{\text{d} T }_{k} + \frac{{C_p}_{\ns}}{T}\right) [C]_{k} + & \nonumber \\
				     & \sum_{k=1}^{-1 + \ns} \biggl( \left(- H_{k} + \frac{W_{k} H_{\ns}}{W_{\ns}}\right) \frac{\partial \dot{\omega} }{\partial T }_{k} + & \nonumber \\
				     & \mathindent[larger] \left(- {C_p}_{k} + \frac{W_{k} {C_p}_{k} }{W_{\ns}} \right) \dot{\omega}_{k}\biggr)\Biggr] & \text{\dconp} \\
\intertext{and similarly:}
\frac{\partial\dot{T}}{\partial{T}}  &= \frac{1}{\sum_{k=1}^{\ns} [C]_{k} {C_v}_{k}} \Biggl[ \frac{\text{d} T }{\text{d} t } \sum_{k=1}^{\ns} \left(- \frac{\text{d} {C_v} }{\text{d} T }_{k} + \frac{{C_v}_{\ns}}{T}\right) [C]_{k} + & \nonumber \\
				     & \sum_{k=1}^{-1 + \ns} \biggl(\left(- U_{k} + \frac{W_{k} U_{\ns}}{W_{\ns}}\right) \frac{\partial \dot{\omega} }{\partial T }_{k} + & \nonumber \\
				     & \mathindent[larger] \left(- {C_v}_{k} + \frac{W_{k} {C_v}_{\ns}}{W_{\ns}}\right) \dot{\omega}_{k}\biggr)\Biggr] & \text{\dconv}
\subsubsection{Molar derivative}
The molar derivative of the energy equation is somewhat easier to work with,

@@

\subsection{Rate of progress derivatives}
\subsubsection{Temperature derivative}
\label{s:dri_dt}
Next, the rate of progress derivatives are evaluated, first with respect to temperature:
\begin{equation}
 \frac{\partial R }{\partial T }_{i} = \frac{\partial}{\partial T}\left({k_f}_{i} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime}_{k,i}}\right)
\end{equation}
The derivative of the forward reaction rate coefficient is:
\begin{equation}
 \label{e:dkf_dt}
 \frac{\text{d} {k_f} }{\text{d} T }_{i} = \frac{{k_f}_{i}}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)
\end{equation}
While evaluating derivatives for the Jacobian, it is important to expand terms---e.g., concentration, production rate, etc.---related to the last species to obtain the correct derivative, in this case using \cref{e:last_spec_conc}:
\begin{equation}
 {R_f} = \left(\left(- \sum_{k=1}^{\ns  - 1} [C]_{k} + \frac{P}{T \Ru}\right)^{\nu^{\prime}_{\ns,i}}\right) {k_f}_{i} \prod_{k=1}^{\ns  - 1} [C]_{k}^{\nu^{\prime}_{k,i}}
\end{equation}
The derivative of a species concentration with respect to temperature is zero for both \conp~and \conv:
\begin{align}
 \frac{\partial [C]_{k}}{\partial T} &= \frac{\partial}{\partial T} \left(\frac{n_k}{V}\right), \quad \left\{k \ne \ns\right\} \nonumber \\
				     &= 0
\end{align}
as the volume is either constant (\conp) or a state variable and a function of time only (\conv).
Hence, the full derivative of the forward rate of progress is:
\begin{equation}
 \label{e:dropfdt_1}
 \frac{\partial {R_f} }{\partial T }_{i} = \frac{\text{d} {k_f} }{\text{d} T }_{i} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime}_{k,i}} - \frac{[C] \nu^{\prime}_{\ns,i}}{T} [C]_{\ns}^{\nu^{\prime}_{\ns,i} - 1} {k_f}_{i} \prod_{k=1}^{\ns  - 1} [C]_{k}^{\nu^{\prime}_{k,i}}
\end{equation}
By defining a temporary variable $S^{\prime}_{l}$ (for species $l = \left\{j, \ns\right\}$):
\begin{equation}
 \label{e:s_temp}
 S^{\prime}_{l} = \nu^{\prime}_{l,i} [C]_{l}^{\nu^{\prime}_{l,i} - 1} \prod_{\substack{1 \leq l \leq l - 1\\l + 1 \leq l \leq \ns}} [C]_{l}^{\nu^{\prime}_{l,i}}
\end{equation}
and using~\cref{e:dkf_dt}, \cref{e:dropfdt_1} can be simplified to:
\begin{equation}
 \label{e:dropf_dt}
 \frac{\partial {R_f} }{\partial T }_{i} = - \frac{[C] S^{\prime}_{\ns}}{T} {k_f}_{i} + \frac{{R_f}_{i}}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)
\end{equation}

Starting from~\cref{e:ropr} and applying the same expansion of the last species' concentration as previously, the temperature derivative of the reverse rate of progress is:
\begin{equation}
 \label{e:droprdt_1}
 \frac{\partial {R_r} }{\partial T }_{i} = \frac{\text{d} {k_r} }{\text{d} T }_{i} \prod_{k=1}^{\ns} [C]_{k}^{\nu^{\prime\prime}_{k,i}} - \frac{[C] \nu^{\prime\prime}_{\ns,i}}{T} [C]_{\ns}^{\nu^{\prime\prime}_{\ns,i} - 1} {k_f}_{i} \prod_{k=1}^{\ns  - 1} [C]_{k}^{\nu^{\prime\prime}_{k,i}}
\end{equation}
Next, \cref{e:kr} is considered to obtain the temperature derivative of the non-explicit reversible reaction rate coefficient:
\begin{align}
 \frac{\text{d} {k_r} }{\text{d} T }_{i} &= \left(- \frac{1}{{K_c}_{i}} \frac{\text{d} {K_c}_{i} }{\text{d} T } \frac{{k_f}_i}{{K_c}_i} + \frac{1}{{K_c}_{i}}\frac{\text{d} {k_f}_{i} }{\text{d} T } \right) \nonumber \\
				         &= \left(- \frac{1}{{K_c}_{i}} \frac{\text{d} {K_c}_{i} }{\text{d} T } + \frac{1}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)\right) {k_r}_{i}
\end{align}
The temperature derivative of the equilibrium constant is:
\begin{equation}
 \frac{\text{d} {K_c}_{i} }{\text{d} T } = {K_c}_{i} \sum_{k=1}^{\ns} \nu_{k,i} \frac{\text{d} B }{\text{d} T }_{k} \;,
\end{equation}
with:
\begin{equation}
 \frac{\text{d} B }{\text{d} T }_{k} = T \left(T \left(\frac{T a_{k,4}}{5} + \frac{a_{k,3}}{4}\right) + \frac{a_{k,2}}{3}\right) + \frac{a_{k,1}}{2} + \frac{1}{T} \left(a_{k,0} - 1 + \frac{a_{k,5}}{T}\right)
\end{equation}
giving:
\begin{equation}
\label{e:dkr_dt}
\frac{\text{d} {k_r} }{\text{d} T }_{i} = \left(- \sum_{k=1}^{\ns} \nu_{k,i} \frac{\text{d} B }{\text{d} T }_{k} + \frac{1}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)\right) {k_r}_{i} \;,
\end{equation}
and using an temporary value $S^{\prime\prime}_{\ns}$---defined analagously to \cref{e:s_temp} for the reverse direction---we obtain:
\begin{equation}
\label{e:dropr_dt}
\frac{\partial {R_r} }{\partial T }_{i} = \left(- \sum_{k=1}^{\ns} \nu_{k,i} \frac{\text{d} B }{\text{d} T }_{k} + \frac{1}{T} \left(\beta_{i} + \frac{{E_{a}}_{i}}{T \Ru}\right)\right) {R_r}_{i} - \frac{[C] S^{\prime\prime}_{\ns}}{T} {k_r}_{i}
\end{equation}
